The illusory promise of the Aligned Rank Transform

### 2025-12-28 appendix_II.qmd
working on appendix

## Absence of spurious effects when $a_2=0$ and $a_{12} = 0${#effects-proof}
Let $x_1$ and $x_2$ be two independent variables, and suppose the conditional distribution of the response $Y | (x_1, x_2)$ depends on these variables only through the linear predictor:
$$
\ell(x_1, x_2) = a_0 + a_1 x_1 + a_2 x_2 + a_{12}x_1 x_2
$$

Assume that the conditional mean can be written as:
$$
\mu(x_1, x_2) = E(Y|x_1, x_2) = h(\ell(x_1, x_2))
$$
where $h$ is a monotone function. This function may coincide with the inverse link function $g^{-1}$ in a GLM, but it may also arise from other frameworks, including transformation models or latent-variable constructions. Thus it does not necessarily need to be smooth or invertible; for example, this assumption is not required for the ordinal example discussed later.

@@

#### Log-normal distributions
Let $x_1$ and $x_2$ be two experimental factors. Suppose the true model of the data is
$$
\begin{aligned}
log Y = a_0 + a_1 x_1 + \varepsilon, \qquad
\varepsilon \sim \mathcal{N}(0, \sigma^2)
\end{aligned} 
$$
so that 
$$Y | (x_1, x_2) \sim Lognormal(a_0 + a_1 x_1, \sigma^2)$$

Since the conditional distribution of $Y$ does not depend on $x_2$, there is no main effect of $x_2$ on the mean, median, or any quantile of the response distribution. In particular, the conditional mean
$$\mu(x_1,x_2) = E(Y|x_1, x_2) = exp(a_0 + a_1 x_1 + \frac{1}{2}\sigma^2)$$
remains identical across all levels of $x_2$.

@@

and the observed response $Y$ is obtained by applying fixed thresholds: 
$$
\begin{aligned}
Y = d\ \ \text{if}\ \ \tau_{d-1} < Y^* \leq \tau_k, 
\qquad d = 1,2 ... D
\end{aligned} 
$$ 
Again, the conditional distribution $Y| (x_1, x_2)$ depends on the independent variables only through $\ell(x_1, x_2)=a_0+a_1 x_1$, and there is no effect of $x_2$ on either the latent or the observed ordinal scale.

@@

<!--
In conclusion, $T$ does not introduce a main effect of $x_2$ or an interaction between $x_1$ and $x_2$ when $a_2=0$ and $a_{12}=0$.

### ART introducing spurious effects 
ART differs fundamentally from simple monotone transformations. Instead of transforming $Y$ directly, ART first constructs aligned responses by **subtracting and adding sample means on the response scale**, for example:

$$Y^{(B)} = Y - \bar{Y}_{ij} + \bar{Y}_{.j} - \bar{Y}_{..}$$
or analogous expressions for interactions, and only then applies ranking. The conditional distribution $Y^{(B)} | (x_1, x_2)$ now depends on $x_2$ through the arithmetic mean components even though $a_2 = 0$ and $a_{12}=0$.

These alignment terms can vary across levels of $x_2$ due to sampling variability alone, and depending on $h$, they can acquire systematic shifts across levels of $x_2$ or combinations of $(x_1, x_2)$, even when the true conditional distribution does not. in particular, the term $\bar{Y}_{.j} - \bar{Y}_{..}$ is constant within each level of $x_2$ and shifts all aligned observations for that level either upward or downward.

After ranking, observations from one level of $x_2$ tend to receive systematically higher or lower ranks. Our results confirm this problem.
-->

>>>>

## On spurious effects when $a_2=0$ and $a_{12} = 0${#effects-proof}
Let $x_1$ and $x_2$ be two independent variables, and suppose the conditional distribution of the response $Y | (x_1, x_2)$ depends on these variables only through the linear predictor:
$$
\ell(x_1, x_2) = a_0 + a_1 x_1 + a_2 x_2 + a_{12}x_1 x_2
$$

Assume that the conditional mean can be written as:
$$
\mu(x_1, x_2) = E[Y|x_1, x_2] = h(\ell(x_1, x_2))
$$
where $h$ is a monotone function. This function may coincide with the inverse link function $g^{-1}$ in a GLM, but it may also arise from other frameworks, including transformation models or latent-variable constructions. Thus it does not necessarily need to be smooth or invertible; for example, this assumption is not required for the ordinal example discussed later.

@@

#### Log-normal distributions
Let $x_1$ and $x_2$ be two experimental factors. Suppose the true model of the data is
$$
\begin{aligned}
log Y = a_0 + a_1 x_1 + \varepsilon, \qquad
\varepsilon \sim \mathcal{N}(0, \sigma^2)
\end{aligned} 
$$
so that 
$$Y | (x_1, x_2) \sim \mathcal{LogN}(a_0 + a_1 x_1, \sigma^2)$$

Since the conditional distribution of $Y$ does not depend on $x_2$, there is no main effect of $x_2$ on the mean, median, or any quantile of the response distribution. In particular, the conditional mean
$$\mu(x_1,x_2) = E[Y|x_1, x_2] = \text{exp}(a_0 + a_1 x_1 + \frac{1}{2}\sigma^2)$$
remains identical across all levels of $x_2$.

@@

and the observed response $Y$ is obtained by applying fixed thresholds: 
$$
\begin{aligned}
Y = d\ \ \text{if}\ \ \tau_{d-1} < Y^* \leq \tau_d, 
\qquad d = 1,2 ... D
\end{aligned} 
$$ 
Again, the conditional distribution $Y| (x_1, x_2)$ depends on the independent variables only through $\ell(x_1, x_2)=a_0+a_1 x_1$, and there is no effect of $x_2$ on either the latent or the observed ordinal scale.

@@

## Calculation of location parameters {#mu}
We recall the linear predictor used to model nonlinear relationships (see the [article](index.html#statistical-modeling)):

$$ 
\ell_{ijk} = \mu + s_k + a_1 x_{1i} + a_2 x_{2j} + a_{12} x_{1i} x_{2j}
$$
where $s_k \sim \mathcal{N}(0, \sigma_s^2)$. Our goal is to determine $\mu$ such as the expected value of the response equals a pre-specified target value $E[Y_{ijk}] = M$.

To simplify notation, we define the fixed-effects component
$$\phi_{ij} = a_1 x_{1i} + a_2 x_{2j} + a_{12} x_{1i} x_{2j}$$

so that

$$ 
\ell_{ijk} = \mu + s_k + \phi_{ij}
$$

**A useful identity.** Let $X$ be a continuous random variable with probability density function $f_X(x)$. For any function $g$, its expectation is defined as

$$
E[g(X)] = \int\limits_{-\infty}^\infty g(x) f_X(x) dx
$$

If $g(x) = e^x$, then

$$
E[e^X] = \int\limits_{-\infty}^\infty e^x f_X(x) dx
$$

When $X \sim \mathcal{N}(0, \sigma^2)$, substituting the normal probability density function yields

$$
E[e^X] = \int\limits_{-\infty}^\infty e^x \frac{1}{\sqrt{2 \pi \sigma^2}} e^{-\frac{x^2}{2\sigma^2}} dx = e^{\frac{1}{2}\sigma^2}
$$ {#eq-useful}

### Log-normal model
When responses are log-normally distributed,

$$
\begin{aligned}
log(Y_{ijk}) = \ell_{ijk} + \varepsilon_{ijk},
\qquad \varepsilon_{ijk} \sim \mathcal{N}(0, \sigma^2)
\end{aligned} 
$$

so that

$$ 
Y_{ijk} = e^{\ell_{ijk} + \varepsilon_{ijk}} = e^{\mu + s_k + e_{ij} + \varepsilon_{ijk}} = e^{\mu} \cdot e^{\phi_{ij}} \cdot e^{s_k} \cdot e^{\varepsilon_{ijk}} 
$$

Taking expectations and using independence of components,

$$
E[Y_{ijk}] = e^\mu \cdot E[e^{\phi_{ij}}] \cdot E[e^{s_k}] \cdot E[e^{\varepsilon_{ijk}}] = M   
$$

Solving for $\mu$ gives

$$
\mu = log\ M - log\ E[e^{\phi_{ij}}] - log\ E[e^{s_k}] - log\ E[e^{\varepsilon_{ijk}}]
$$

Because $s_k \sim \mathcal{N}(0, \sigma_s^2)$ and $\epsilon_{ijk} \sim \mathcal{N}(0, \sigma^2)$, @eq-useful implies

$$
\begin{aligned}
E[e^{s_k}] = e^{\frac{1}{2}\sigma_s^2},
\qquad E[e^{\varepsilon_{ijk}}] = e^{\frac{1}{2}\sigma^2}
\end{aligned} 
$$

Furthermore, since $\phi_{ij}$ is deterministic,

$$E[e^{\phi_{ij}}] = \frac{1}{m_1 m_2} \sum\limits_{i=1}^{m_1}\sum\limits_{j=1}^{m_2} e^{\phi_{ij}}$$
where $m_1$ and $m_2$ are the number of levels of $X_1$ and $X_2$, respectively.

Therefore,

$$
\mu = log\ M - log(\frac{1}{m_1 m_2} \sum\limits_{i=1}^{m_1}\sum\limits_{j=1}^{m_2} e^{\phi_{ij}}) - \frac{1}{2}\sigma_s^2 - \frac{1}{2}\sigma^2
$$ {#eq-mu-lognornal}

### Exponential model
For exponential responses with a log link,

$$
Y_{ijk} \sim \text{Exp}(\lambda_{ijk} = e^{-\ell_{ijk}})
$$

The expected value is

$$
E[Y_{ijk}] = E[1/\lambda_{ijk}] = E[e^{\ell_{ijk}}]  =  e^{\mu} \cdot E[e^{\phi_{ij}}] \cdot E[e^{s_k}]
$$

which, as before, it leads to the solution

$$
\mu = log\ M - log(\frac{1}{m_1 m_2} \sum\limits_{i=1}^{m_1}\sum\limits_{j=1}^{m_2} e^{\phi_{ij}}) - \frac{1}{2}\sigma_s^2
$$ {#eq-mu-exp}

### Poisson model
For Poisson responses with a log link,
$$
Y_{ijk} \sim \mathcal{Pois}(\lambda_{ijk} = e^{\ell_{ijk}})
$$

The expected value is

$$
E[Y_{ijk}] = E[\lambda_{ijk}] = E[e^{\ell_{ijk}}]  =  e^{\mu} \cdot E[e^{\phi_{ij}}] \cdot E[e^{s_k}]
$$
which leads to the same solution as for the exponential distribution:
$$
\mu = log\ M - log(\frac{1}{m_1 m_2} \sum\limits_{i=1}^{m_1}\sum\limits_{j=1}^{m_2} e^{\phi_{ij}}) - \frac{1}{2}\sigma_s^2
$$ {#eq-mu-poisson}

### Binomial model
For binomial responses with a logit link,

$$
Y_{ijk} \sim \mathcal{B}(\kappa, p_{ijk} = \frac{1}{1 + e^{-\ell_{ijk}}})
$$

The expected value is now

$$E[Y_{ijk}] = \kappa \cdot E[p_{ijk}] = 
\frac{\kappa}{m_1 m_2} \sum\limits_{i=1}^{m_1}\sum\limits_{j=1}^{m_2}
\int_{-\infty}^{\infty}
\frac{1}{1 + e^{-(\mu + \phi_{ij} + s)}}
\cdot
\frac{1}{\sqrt{2\pi}\sigma_s}
e^{-\frac{s^2}{2\sigma_s^2}}\, ds
= M
$$ {#eq-mu-binomial}
because we cannot separate the sums from the integral. As no close form exists to analytically derive $\mu$, we create the function

$$F(\mu) = \frac{\kappa}{m_1 m_2} \sum\limits_{i=1}^{m_1}\sum\limits_{j=1}^{m_2}
\int_{-\infty}^{\infty}
\frac{1}{1 + e^{-(\mu + \phi_{ij} + s)}}
\cdot
\frac{1}{\sqrt{2\pi}\sigma_s}
e^{-\frac{s^2}{2\sigma_s^2}}\, ds
- M$$

and numerically find its root using R's ``uniroot()`` function.

### Cauchy model
Since the Cauchy distribution has no finite mean, we target the median instead:

$$\text{Median}(Y_{ijk}) = \ell_{ijk} = \mu + \phi_{ij} = M$$

Because our coding is centered, $\sum\limits_{i=1}^{m_1}\sum\limits_{j=1}^{m_2} \phi_{ij} = 0$. Therefore,

$$
\mu  = M
$$

### 2025-12-28 sandbox.qmd
working on appendix

```{=html}
<style>
.math.inline .MathJax  {
  font-size: 105% !important;
}
</style>
```
### Type I errors under unequal variances
Many statistical procedures assume equal variances across all levels of each independent variable, or more strictly, across all possible pairs of factor levels (the latter is known as the sphericity assumption in repeated-measures ANOVA). Nonparametric tests are often mistakenly assumed to be free of such assumptions; however, as discussed earlier, this is generally incorrect.

Our next experiment evaluates the behavior of the four methods under conditions of unequal variances, focusing on the normal distribution and four ordinal distributions. For ordinal outcomes, we control variances at the level of the latent variable $Y^*$.

We examine three two-factor designs: (i) a $4 \times 3$ within-subjects design; (ii) a $2 \times 3$ between-subjects design; and (iii) a $2 \times 4$ mixed design. We set all effects to zero and vary the ratio $r_{sd}$ of standard deviations between the levels of the first factor $X_1$. @fig-unequal shows the five ratios tested when $X_1$ has two levels. When $X_1$ has additional levels (as in the $4 \times 3$ design), their standard deviations are randomly drawn from a range bounded by the standard deviations of the first two levels.

::: {#fig-unequal}
```{r, echo=FALSE, message=FALSE, fig.height=1.5, fig.width = 9, warning=FALSE}
source("distributions_plots.R")
plot_unequal_variances()
```
Varying the ratio $r_{sd}$ of standard deviations across the levels of $X_1$. The plots show examples of population distributions for factors with two categorical levels.    
:::

**Main effects**. We begin by investigating how the four methods detect main effects of $X_1$. Interpreting results in this setting is challenging because the underlying populations differ in variance while sharing the same means (and medians). For ordinal scales, differences in variance may also manifest as differences in observed means [@Liddell:2018]. In this experiment, we define the null hypothesis in terms of mean differences on the latent variable, but we emphasize that the results should be interpreted with caution.

@fig-unequal-main-1 presents the results. Under normal distributions, and depending on the experimental design, the different methods either slightly inflate or slightly deflate Type I error rates as $r_{sd}$ increases, with no method clearly outperforming the others. For ordinal scales, larger values of $r_{sd}$ primarily affect conditions with flexible thresholds. This effect is particularly pronounced for PAR, although all methods are affected.

::: {#fig-unequal-main-1}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "Type_I_heteroscedacity"
distributions = c("norm", "likert_5", "likert_5_flex", "likert_7", "likert_7_flex")
dnames = c("Normal", "Equidistant (5 levels)", "Flexible (5 levels)", "Equidistant (7 levels)", "Flexible (7 levels)")

df <- read_data(prefix, alpha = 0.05, effectType = 1, distributions)
df <- reshape_by_design(df, dnames, effectvars = c("sdratio", "effectX1","effectX2","effectX1X2"))
plotly_error_by_design_2(df, xlab = "max ratio of standard deviations between levels of X1", var = "rateX1", xvar = "sdratio", max = 41)
```
Type I error rates ($\alpha = .05$) for the **main effect of $X_1$** as the ratio of standard deviations $r_{sd}$ for $X_1$ increases ($n=20$). When $r_{sd} > 1$, the underlying populations are no longer identical, and different interpretations of the null hypothesis may lead to different outcomes.
:::

We also examine the influence of unequal variances across the levels of $X_1$ on the Type I error rate for $X_2$, as shown in @fig-unequal-main-2. In this case, there is no ambiguity in the interpretation of effects because the populations defined by $X_2$ are identical.

The Type I error rates of PAR, RNK, and INT appear largely unaffected by increases in $r_{sd}$. In contrast, ART yields high error rates, surpassing acceptable levels even under normal distributions (see, for example, the $2 \times 4$ mixed design). Thus, unequal variances pose a more serious problem for ART than for the other methods, as the method's sensitivity propagates across factors. This provides further evidence that ART's alignment procedure leads to confounding across effects.

::: {#fig-unequal-main-2}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
plotly_error_by_design_2(df, xlab = "max ratio of standard deviations between levels of X1", var = "rateX2", xvar = "sdratio", max = 41)
```
Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as the ratio of standard deviations $r_{sd}$ across the levels of $X_1$ increases ($n=20$)
:::

**Interaction effects**. Finally, we evaluate Type I error rates for the interaction, as shown in @fig-unequal-interaction. ART exhibits a similar pattern as before, with error rates increasing as $r_{sd}$ grows. We also observe some inflation of Type I error rates for the other three methods under the $4 \times 3$ within-subjects design, albeit to a significantly lesser degree.

::: {#fig-unequal-interaction}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
plotly_error_by_design_2(df, xlab = "max ratio of standard deviations between levels of X1", var = "rateX1X2", xvar = "sdratio", max = 42)
```
Type I error rates ($\alpha = .05$) for the **interaction effect $X_1 \times X_2$** as the ratio of standard deviations $r_{sd}$ across the levels of $X_1$ increases ($n=20$)
:::

>>>>

```{=html}
<style>
.math.inline .MathJax  {
  font-size: 105% !important;
}
</style>
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(tidyverse)

source("utils-data-reading.R")
source("utils-plotlying.R")
```

```{=html}
<style>
.math.inline .MathJax  {
  font-size: 105% !important;
}
</style>
```

We present complementary results and new experiments that investigate additional scenarios. We also compare INT and RNK with other nonparametric methods. Unless explicitly mentioned in each section, we follow the experimental methodology presented in the main article. At the end of each section, we summarize our conclusions.

## Generalizations of nonparametric tests {#generalized}

Finally, we evaluate the generalizations of nonparametric tests recommended by Lüpsen [-@luepsen:2018; -@luepsen:2023] as implemented in his *np.anova* function [@luepsen_R]. Specifically, we examine the generalization of the van der Waerden test (VDW) and the generalization of the Kruskal-Wallis and Friedman tests (KWF).

These implementations require random slopes to be included in the error term of the model. Concretely, we use ``Error(Subject/(X1*X2))`` for the two-factor within-participants design and ``Error(Subject/X2)`` for the mixed design. We use the same modeling approach for all other methods to ensure comparability.

**Type I error rates: Main effects**. @fig-vdWaerden-designs-main shows the Type I error rates for the main effect of $X_2$. While all methods perform well under the within-subjects and mixed designs, the error rates of VDW and KWF drop sharply as the effect of $X_1$ increases under the between-subjects design. As shown below, this behavior reflects a severe loss of power for these two methods in these conditions.

::: {#fig-vdWaerden-designs-main}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "Type_I_vdWaerden"
alpha = 0.05

distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert")
dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
methods = c("RNK", "INT", "VDW", "KWF")
palette = c("#E69F00", "#009E73", "#FF70AB", "#888888")

df <- read_data(prefix, alpha, effectType = 1, distributions, methods = methods)
df <- reshape_by_design(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))

plotly_error_by_design_2(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 8.1, nticks=8, cbPalette = palette)
```
Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_1$ of the main effect of $X_1$ ($n = 20$)
:::

**Type I error rates: Interactions**. @fig-vdWaerden-designs-interaction-1 presents the Type I error rates for the interaction in the presence of a single main effect. Again, the error rates of VDW and KWF decrease rapidly, now in both the between-subjects and mixed designs. We also observe inflation of the error rates of RNK and INT for large effects under discrete distributions.

::: {#fig-vdWaerden-designs-interaction-1}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- read_data(prefix, alpha, effectType = 2, distributions, methods = methods)
df <- reshape_by_design(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))

plotly_error_by_design_2(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX2", max = 8.1, nticks=8, cbPalette = palette)
```
Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitude $a_2$ of the main effect of $X_2$ ($n = 20$) 
:::

@fig-vdWaerden-designs-interaction-2 shows the results when both $a_1$ and $a_2$ increase. In these settings, all four methods struggle as effect magnitudes grow, though their behavior varies across distributions and experimental designs. Depending on the configuration, the methods either inflate positive rates or deflate them below nominal levels. INT appears to be the most stable method for continuous distributions, but its positive rates increase more rapidly under discrete distributions.

::: {#fig-vdWaerden-designs-interaction-2}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- read_data(prefix, alpha, effectType = 0, distributions, methods = methods)
df <- reshape_by_design(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotly_error_by_design_2(df, xlab = "magnitude of main effects", ytitle = 'Positives (%)', var = "rateX1X2", xvar = "effectX1", max = 104, nticks=8, cbPalette = palette)
```
Positive rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$ ($n = 20$). *Note: These values correspond to Type I error rates under the null hypothesis $a_{12} = 0$; alternative definitions of the null hypothesis may lead to different results.* 
:::

**Power: Main effects**. @fig-vdWaerden-power-main-1 presents power results for detecting the main effect of $X_1$. Although differences are generally small, INT emerges as the most powerful method in most configurations. In several cases (e.g., log-normal distributions in the within-subjects and mixed designs), the power of VDW and KWF lags behind that of RNK and INT.

::: {#fig-vdWaerden-power-main-1}
<div id="plot-power-vd.1">
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "Power_vdWaerden"
distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert")
dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")

data <- read_data(prefix, alpha = .05, effectType = 3, distributions, methods = methods)
df <- reshape_by_design(data, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotly_error_by_design_2(df, xlab = "magnitude of main effect", var = "rateX1", xvar = "effectX1", max = 104,  ytitle = 'Power (%)', cbPalette = palette)
```
</div>

<div id="plot-power-vd.2">
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- data %>% arrange(design,family,effectX1,rateX1)  %>% group_by(design,family,effectX1) %>% mutate(rank = rank(rateX1))
df <- as.data.frame(df) %>% reshape_by_design(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotly_power_by_design(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX1", xvar = "effectX1", max = 4.2, ytitle = 'Power - ranking', cbPalette = palette)
```
</div> <style> .hidden { display: none; } </style> <script> document.addEventListener("DOMContentLoaded", function() { document.getElementById("plot-power-vd.2").classList.add("hidden"); }); </script>

<button id="btnvd1" style="background-color: #f0f0f0; border: 1px solid #ccc; color: #333; font-size: 0.9em;
    padding: 4px 10px; border-radius: 6px; cursor: pointer; float: right;"
  onclick="const plot = document.getElementById('plot-power-vd.1');
  plot.classList.toggle('hidden');
  document.getElementById('plot-power-vd.2').classList.toggle('hidden');
  btnvd1.innerText = plot.classList.contains('hidden') ? 'show percent' : 'show ranking';
">
  show ranking
</button>

Power ($\alpha = .05$) for detecting the **main effect of $X_1$** as a function of the magnitude of effect $a_1$ ($n=20$).
:::

We also report power results for the second factor $X_2$, which confirm that VDW and KWF offer no clear advantage, particularly when compared with INT.

::: {#fig-vdWaerden-power-main-2}
<div id="plot-power-vd.3">
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "Power_vdWaerden"
distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert")
dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")

data <- read_data(prefix, alpha = .05, effectType = 4, distributions, methods = methods)
df <- reshape_by_design(data, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotly_error_by_design_2(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX2", max = 104,  ytitle = 'Power (%)', cbPalette = palette)
```
</div>

<div id="plot-power-vd.4">
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- data %>% arrange(design,family,effectX2,rateX2)  %>% group_by(design,family,effectX2) %>% mutate(rank = rank(rateX2))
df <- as.data.frame(df) %>% reshape_by_design(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotly_power_by_design(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX2", xvar = "effectX2", max = 4.2, ytitle = 'Power - ranking', cbPalette = palette)
```
</div> <style> .hidden { display: none; } </style> <script> document.addEventListener("DOMContentLoaded", function() { document.getElementById("plot-power-vd.4").classList.add("hidden"); }); </script>

<button id="btnvd2" style="background-color: #f0f0f0; border: 1px solid #ccc; color: #333; font-size: 0.9em;
    padding: 4px 10px; border-radius: 6px; cursor: pointer; float: right;"
  onclick="const plot = document.getElementById('plot-power-vd.3');
  plot.classList.toggle('hidden');
  document.getElementById('plot-power-vd.4').classList.toggle('hidden');
  btnvd2.innerText = plot.classList.contains('hidden') ? 'show percent' : 'show ranking';
">
  show ranking
</button>

Power ($\alpha = .05$) for detecting the **main effect of $X_2$** as a function of the magnitude of effect $a_2$ ($n=20$).
:::

**Power: Interactions**. We also report power results for detecting interaction effects in @fig-vdWaerden-power-interaction. These results further confirm the overall advantage of INT across all designs.

::: {#fig-vdWaerden-power-interaction}
<div id="plot-power-vd.5">
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "Power_vdWaerden"
distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert")
dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")

data <- read_data(prefix, alpha = .05, effectType = 5, distributions, methods = methods)
df <- reshape_by_design(data, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotly_error_by_design_2(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1X2", max = 104,  ytitle = 'Power (%)', cbPalette = palette)
```
</div>

<div id="plot-power-vd.6">
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- data %>% arrange(design,family,effectX1X2,rateX1X2)  %>% group_by(design,family,effectX1X2) %>% mutate(rank = rank(rateX1X2))
df <- as.data.frame(df) %>% reshape_by_design(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotly_power_by_design(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX1X2", xvar = "effectX1X2", max = 4.2, ytitle = 'Power - ranking', cbPalette = palette)
```
</div> <style> .hidden { display: none; } </style> <script> document.addEventListener("DOMContentLoaded", function() { document.getElementById("plot-power-vd.6").classList.add("hidden"); }); </script>

<button id="btnvd3" style="background-color: #f0f0f0; border: 1px solid #ccc; color: #333; font-size: 0.9em;
    padding: 4px 10px; border-radius: 6px; cursor: pointer; float: right;"
  onclick="const plot = document.getElementById('plot-power-vd.5');
  plot.classList.toggle('hidden');
  document.getElementById('plot-power-vd.6').classList.toggle('hidden');
  btnvd3.innerText = plot.classList.contains('hidden') ? 'show percent' : 'show ranking';
">
  show ranking
</button>

Power ($\alpha = .05$) for detecting the **interaction effect $X_1 \times X_2$** as a function of the magnitude of effect $a_{12}$ ($n=20$).
:::

### 2025-12-28 text/50-ExperimentalMethod.qmd
working on appendix

 - $\mu$ is the grand mean

@@

**Magnitude of effects.** We control the magnitude of effects by varying $a_1$, $a_2$, and, for some experiments (those evaluating power for interactions), $a_{12}$. @fig-effects-normal presents normal distributions ($\sigma^2 = 1$) generated across the range of values of $a_1$ used to evaluate Type I error rates. In these examples, we illustrate effects for a factor with three categorical levels and, for clarity, disregard between-subject variability ($\sigma_k^2 = 0$).

@@

**Grand mean and variance parameters.** We select the mean term $\mu$ to match a target location of the responses. @tbl-locations reports the target location for each distribution, which is typically the overall mean (or the median in the case of the Cauchy distribution). We describe how $\mu$ is derived from these parameters in [Appendix II](appendix_II.html#mu).

@@

::: {#fig-effects-distributions}
```{r, echo=FALSE, message=FALSE, fig.height=1.2, fig.width = 9, warning=FALSE}
plot_effects_distr()
```
Different ratio distributions generated with $a_1 = 2$ when $X_1$ has three categorical levels (here, $\sigma_k^2 = 0$).
:::

In practice, we randomly choose the standard deviation $\sigma_k^2$ --- which determines between-subject variability --- from the interval $[0.1, 0.5]$, drawing each value from a uniform distribution.

>>>>

 - $\mu$ is the intercept

@@

**Magnitude of effects.** We control the magnitude of effects by varying $a_1$, $a_2$, and, for some experiments (those evaluating power for interactions), $a_{12}$. @fig-effects-normal presents normal distributions ($\sigma^2 = 1$) generated across the range of values of $a_1$ used to evaluate Type I error rates. In these examples, we illustrate effects for a factor with three categorical levels and, for clarity, disregard between-subject variability ($\sigma_s^2 = 0$).

@@

**Intercept and variance parameters.** Fixing the intercept $\mu$ directly in a nonlinear model generally leads to arbitrary population-level response distributions, because the marginal mean depends on the link function, the response distribution, the variance of the random effects, and the magnitude of the fixed effects. We therefore calibrate the intercept $\mu$ so that the induced location of the responses matches a pre-specified target value. @tbl-locations reports the target location for each distribution, which is typically the overall mean in all cases except for the Cauchy distribution, where the median is used. We detail how $\mu$ is analytically derived from these target values in [Appendix II](appendix_II.html#mu).

@@

::: {#fig-effects-distributions}
```{r, echo=FALSE, message=FALSE, fig.height=1.2, fig.width = 9, warning=FALSE}
plot_effects_distr()
```
Different ratio distributions generated with $a_1 = 2$ when $X_1$ has three categorical levels (here, $\sigma_s^2 = 0$).
:::

In practice, we randomly choose the standard deviation $\sigma_s^2$ --- which determines between-subject variability --- from the interval $[0.1, 0.5]$, drawing each value from a uniform distribution.

### 2025-12-28 text/63-Results-UnequalVariance.qmd
working on appendix

**Main effects**. We begin by investigating how the four methods detect main effects of $X_1$. Interpreting results in this setting is challenging because the underlying populations differ in variance while sharing the same means (and medians). For ordinal scales, differences in variance may also manifest as differences in observed means [@Liddell:2018]. In this experiment, we define the null hypothesis in terms of mean differences on the latent variable, but we emphasize that the results should be interpreted with caution.

@fig-unequal-main-1 presents the results. Under normal distributions, and depending on the experimental design, the different methods either slightly inflate or slightly deflate Type I error rates as $r_{sd}$ increases, with no method clearly outperforming the others. For ordinal scales, larger values of $r_{sd}$ primarily affect conditions with flexible thresholds. This effect is particularly pronounced for PAR, although all methods are affected.

@@

df <- read_data(prefix, alpha = 0.05, effectType = 1, distributions)
df <- reshape_by_design(df, dnames, effectvars = c("sdratio", "effectX1","effectX2","effectX1X2"))
plotly_error_by_design_2(df, xlab = "max ratio of standard deviations between levels of X1", var = "rateX1", xvar = "sdratio", max = 41)
```
Type I error rates ($\alpha = .05$) for the **main effect of $X_1$** as the ratio of standard deviations $r_{sd}$ for $X_1$ increases ($n=20$). When $r_{sd} > 1$, the underlying populations are no longer identical, and different interpretations of the null hypothesis may lead to different outcomes.
:::

>>>>

**Main effects**. We begin by investigating how the four methods detect main effects of $X_1$. Interpreting results in this setting is challenging because the underlying populations differ in variance while sharing the same means (and medians). For ordinal scales, differences in variance may also manifest as differences in observed means [@Liddell:2018]. To highlight these interpretation issues,we therefor label the observed rates of positives as *Positives (%)*. These rates can nonetheless be interpreted as Type I error rates if the null hypothesis is defined in terms of mean differences on the latent variable.

@fig-unequal-main-1 presents the results. Under normal distributions, and depending on the experimental design, the different methods either slightly inflate or slightly deflate the positive rates as $r_{sd}$ increases, with no method clearly outperforming the others. For ordinal scales, larger values of $r_{sd}$ primarily affect conditions with flexible thresholds. This effect is particularly pronounced for PAR, although all methods are affected.

@@

df <- read_data(prefix, alpha = 0.05, effectType = 1, distributions)
df <- reshape_by_design(df, dnames, effectvars = c("sdratio", "effectX1","effectX2","effectX1X2"))
plotly_error_by_design_2(df, xlab = "max ratio of standard deviations between levels of X1", ytitle = 'Positives (%)', var = "rateX1", xvar = "sdratio", max = 41)
```
Positive rates ($\alpha = .05$) for the **main effect of $X_1$** as the ratio of standard deviations $r_{sd}$ for $X_1$ increases ($n=20$). *Note: When $r_{sd} > 1$, the underlying populations are no longer identical, and different interpretations of the null hypothesis may lead to different outcomes.*
:::
